---
title: "Data Visualisation in R (Base, Lattice, ggplot2)"
author:  | 
  | Arash Ardalan
  | a.ardalan07@gmail.com
date: "24/10/2022"
output:
  beamer_presentation: default
  ioslides_presentation: default
  extra_dependencies:
  - bbm
  - threeparttable
  - xcolor
  - colortbl
  - tcolorbox
  - bmatrix
  slidy_presentation: default
urlcolor: blue
linkcolor: red
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Overview of R graphics
1. Standard `graphics` in R
  + The `graphics` package, as the \textcolor{blue}{base graphics system},
provides a complete set of functions for creating a wide variety of plots.

## grid system:

2. Trellis & `lattice` graphics
  
  + The techniques were given the name Trellis because: 
 usually results are in a \textcolor{blue}{rectangular array of plots}.
 
 3. Grammar of Graphics `ggplot2` 
  + A powerful approach, based on the \textcolor{blue}{“Grammar of Graphics”}
  
  


# Expectation Setting
#### I am assuming the following:
- You are experienced with R coding - not an expert, but you can hack.
- You have some statistical knowledge (e.g., what is a histogram or Boxplot).

#### This is a quick intro to data visualization with R:
- I will gloss over a lot of things .
- More in-depth coverage is available via resources I will mention at the end.
- My goal is to make you excited about graphics in R.
  - [GitHub URL:][https://github.com/ArashArd/Graphics-in-R]

#### Prerequisites
- To follow along you will need the following:
  - [R](https://cran.r-project.org/)
  - [RStudio](https://www.rstudio.com/)
  
  
# Types of more common use Statistical Graphs 
            Types of Data (Variables)
 
### Numeical Variables 
- Estimation and shape of Distribution
  - Histogram and Density plots
  - Box-plot
- Comparisons  
  - Box-plots \& Violin plots
- Associations and finding the structure between variables 
  - Scatter plots
  - Time series plots
- Other Advanced Graphic Tools

### Categorical Variables
- Count and Percentages
  - Bar charts and Pie charts

### Explore Graphics 
  

# Data Sets Which I am using here
\small 
## Titanic Data set

RMS Titanic was a British passenger liner which sank in the North Atlantic Ocean on 15 April 1912 after striking an iceberg. Voyage:  from UK to US.

\scalebox{0.4}{
```{r hmisc-table1, echo=FALSE, message=FALSE, results='asis'}
library(readxl)
library(Hmisc)
Titanic <- read_excel("C:/Users/Arash/OneDrive - Royal Holloway University of London/Desktop/GWR/Titanic.xlsx")
latex(head(Titanic, 5), file='', table.env=FALSE, center='none')
```
}


## `USArrests` Data set
Violent Crime Rates by US State


\scalebox{0.55}{
```{r hmisc-table2, echo=FALSE, message=FALSE, results='asis'}

latex(head(USArrests, 5), file='', table.env=FALSE, center='none')
```
}

## 'mtcars' Data set
Motor Trend Car Road Tests

\scalebox{0.55}{
```{r hmisc-table3, echo=FALSE, message=FALSE, results='asis'}

latex(head(mtcars, 5), file='', table.env=FALSE, center='none')
```
}


# Types of Graphs in R 
 \begin{itemize}
 \item High Level\\
Some examples of high level graphics functions are:
\begin{itemize}
\item  Pie Charts, Bar Charts and Histograms
\item Box-and-Whisker Plots
\item Scatter plots
\item Time Series Plots
\item  Surface Plots
\item Other Advanced Plots
\end{itemize}
 \item Low Level\\
The high-level graphics facilities in R are built on a set of
flexible low-level ones. The low level facilities include:
\begin{itemize}
\item Page Layout
\item Setup of Plotting Coordinates
\item Drawing Points and Lines
\item Drawing Polygons and Rectangles
\item Color Management
\end{itemize}
\end{itemize}

# `graphics` package: base graphics system  
\begin{enumerate}
\item \textbf{\textcolor{purple}{plot}}\\
The \textbf{`plot`} function produces an entire graph with a single
function call.\\
It can be useful to access graphics functionality at \textcolor{blue}{a lower
level} so that it is possible to create graphs in a much more
flexible way.
 \item \textbf{\textcolor{purple}{plot.new}}
 \item \textbf{\textcolor{purple}{plot.window}}
\begin{itemize}
  \item
The functions \textbf{plot.new} and \textbf{plot.window} are the functions
which make it possible to work in this low-level way.
\item
\textbf{plot.new} is used to \textcolor{purple}{begin a new plot}, 
\item \textbf{plot.window} is used to \textcolor{purple}{set up coordinate systems}.
\item
Neither function does any actual drawing.
\end{itemize}

\end{enumerate}

# Graphical commands
\small
#### High level graphical commands create the plot

commands |  Type of plots
---------- |-----------
 plot()     |        Scatter plot, and general plotting
 hist()            | Histogram
  barplot()        | Barplot
  boxplot()        | Boxplot
 pairs()           | Plots for multivariate data (Matrix scatter plots)
 qqnorm()          | Normal probability plot

####  Low level graphical commands add to the plot
\scriptsize
commands |  Description
---------- |-----------
 points()   | Add points
 lines()    | Add lines
 rect()     | Add rectangle 
 text()     | Add text
 abline()   | Add lines
 arrows()   | Add arrows
 segments() | add line segment
 legend()   | Add legend


# A Simple High-Level Plot
\small
```{r, echo = TRUE ,fig.width= 8, fig.height = 4}
x = 1:30 ; y = rnorm(30) + x/5 # Generating x and y: numeric

plot(x = x, y = y, xlim = range(x), ylim = range(y), 
     xlab = "X Coordinates", ylab = "Y Coordinates", 
     main = "A Filled Plot Region", type = 'n')
lines(x, y, lwd = 3, col = "blue")
points(x, y, pch = 19, col = "red")
```

# `plot` function: all in one
The `type` argument of the plot function

commands |  Description
---------- |-----------
 plot(\dots, type = "p") | points
 plot(\dots, type = "l") | lines
 plot(\dots, type = "b") | points connected by lines
 plot(\dots, type = "o") | lines are over the points
 plot(\dots, type = "h") | vertical lines
 plot(\dots, type = "s") | steps
 plot(\dots, type = "n") | No plotting
 Type of the line
\begin{itemize}
\item plot(\dots, type = "l", lty = 1) \# solid (lty: line type)
\item plot(\dots, type = "l", lty = 2) \# dashed
\item plot(\dots, type = "l", lty = 3) \# dotted
\item plot(\dots, type = "l", lty = 4) \# dotdash
\item plot(\dots, type = "l", lty = 5) \# longdash
\item plot(\dots, type = "l", lty = 6) \# twodash
\end{itemize}



# Point Characters
\footnotesize
```{r , echo = TRUE,  fig.height = 5}
plot.new(); plot.window(xlim = c(4, 26), ylim = c(3, 26) )

 r = rep(1:5 * 5, each = 5); t = rep(5:1 * 5, 5)
 points(r, t, 
        pch = 1:25,     # Symbol
        cex = 3,        # Size of the symbol
        col = "blue",   # Border color of the symbol
        bg  = "green")  # Background color of the symbol
        
 text(r - 1.5, t, 1:25)
```


# Drawing a Boxplot : Step by Step, using Low-level graphics


\tiny
```{r , echo = TRUE,  fig.height = 2} 
par(mar = c(2, 2.5, 0, 0)) # Setting the Margins 
mydata = rnorm(n = 50, mean = 1, sd = 4); x = mydata; n = length(mydata) # Generating 50 data from Normal   

Minx = min(x); Maxx = max(x); Quarts = quantile(x, c(0.25, 0.5, 0.75)) # Finding the min, max and Quartiles
SM = c(Minx, Quarts, Maxx); names(SM) = c('Min', paste('Q', 1:3, sep = ''), 'Max') # Summary of data in 5 points
SM
plot.new()   
plot.window(xlim = c(Minx - 0.15 * sd(x), Maxx + 0.15 * sd(x)) , ylim = c(0.2, 2) )

rect(xleft = SM[2], ybottom = 0.5, xright = SM[4], ytop = 1.5, lwd = 4 , col = 3, border = "gray60")

segments(x0 = SM[1], y0 = 1, x1 = SM[2], y1 = 1, lty = 'dashed')
segments(x0 = SM[4], y0 = 1, x1 = SM[5], y1 = 1, lty = 'dashed')

segments(SM[1], 0.75, SM[1], 1.25, lwd = 2)
segments(SM[5], 0.75, SM[5], 1.25, lwd = 2)
segments(SM[3], 0.5, SM[3],1.5, lwd = 2, col = "gray30" )
axis(1, at = round(SM,2), labels = names(SM)); axis(2, las = 2)

points(x = mydata, y = runif(n, 0.95, 1.05), col = 2, pch = 19)
```

# A Histogram : Step by Step, using Low-level graphics   


\tiny
```{r , echo = TRUE,  fig.height = 2} 
par(mar = c(2, 2.5, 2, 0))

k = 5                      # number of classes or bars
widthx = (Maxx - Minx)/k   # Setting the bin 

cut_ps = seq(Minx, Maxx, by = widthx)  # Setting the cut-points 
grx    = cut(x, cut_ps)                # classified the data in groups
(TF    = table(grx))                   # Tabulate the grouped 
 
plot.new()
plot.window(xlim = c(Minx - 0.15 * sd(x), Maxx + 0.15 * sd(x)) , ylim = c(0, max(TF)) )
Hist_col = 'green'
for(i in 1:length(TF))
polygon(x = rep(cut_ps[i:(i+1)], each = 2), y = c(0, TF[i], TF[i], 0), col = Hist_col)
 
axis(1, at = round(cut_ps, 2)); axis(2, at = TF, las = 2)
 
title(main ='the Main title', sub ='the subtitle', xlab = 'xlab', ylab = 'ylab')
```

# Control graphical parameters by `par` function 1
\small
```{r, background = 2 , echo =TRUE, fig.height= 3}

par(mfrow = c(2, 2)) # a 2*2 array of figures and combine plots
par(mar = c(3, 4, 0.3, 2.1)) # The size of margins
par(oma = c(3, 4, 2, 4))  # The size of outer margins
x = c(-1, 1)
plot(x, x, type = "n"); text(0, 0, 'Fig1') 
plot(x, x, type = "n"); text(0, 0, 'Fig2')
plot(x, x, type = "n"); text(0, 0, 'Fig3')
plot(x,x, type = "n");  text(0, 0, 'Fig4'); 
box("figure", col = 2) 
title(main = "Plots with Margins Trimmed", outer = TRUE)
box(which = 'outer', col = 'blue', lty = "dashed")
```



# Control graphical parameters by `par` function 2
\small
```{r, background = 2 , echo =TRUE, fig.height= 4}

par(mfrow = c(2, 2)) # a 2*2 array of figures
par(mar = c(5.1, 4.1, 0.1, 2.1)) # The size of margins
par(oma = c(0, 0, 4, 0))  # The size of outer margins

x = rnorm(10)
plot(x, type = "p"); plot(x, type = "l")
plot(x, type = "b"); plot(x, type = "o")
title(main = "Plots with Margins Trimmed", outer = TRUE)
```

# Control graphical parameters by `par` function 3
\scriptsize
```{r, background = 2 , echo =TRUE, fig.height= 4}
x = rnorm(n = 150, mean = 0, sd = 0.8)

par(mfcol = c(2, 2))  # Two rows, two columns
par(mar = c(2.1, 2.1, 2.1, 0.1))
par(bg = "aliceblue") # Aliceblue background color

hist(x, probability = TRUE, col = 'blue', main = "Fig1")    # Top left
lines(density(x), col = "red", lwd = 2)                     # Add a line graph
boxplot(x, col = 'green', main = "Fig2", horizontal = TRUE) # Bottom left      
barplot(table(round(x)), col = 'orange', main = "Fig3")     # Top right
pie(table(round(x)), main = "Fig4")                         # Bottom right
```



#  Plots and coordinates 
\footnotesize
::: columns 
:::: column
```{r, echo=TRUE}
# Cartesian coordinates
x = 1:30
y = rnorm(30) + x/5
plot(x, y, type = "n")
# add 3 lines 
lines(x, y, lwd = 7)
lines(x, y, lwd = 4, 
      col = "green4")
lines(x, y, lwd = 1, 
      col = "white")
```
::::

:::: column
```{r, echo = TRUE}
# polar coordinate
th = seq(0, 2*pi, length = 73)[1:72]
x = cos(th); y = sin(th)
plot.new()
plot.window(xlim = c(-1,1), 
            ylim = c(-1,1),
            asp = 1)
polygon(x, y, 
        col = "darkseagreen")
```
:::: 

:::::
# More on `par` function
\small
R graphics are controlled by use of the par function.
`par` makes it possible to control low-level graphics by
querying and setting a large set of graphical parameters.
Graphical parameters control many features such as:

- the layout of figures on the device
- the size of the margins around plots
- the colours, sizes and typefaces of text
- the colour and texture of lines
- the style of axis to be used
- the orientation of axis labels

\tiny
```{r, echo = TRUE}
names(par())
```
# Background color
\tiny
```{r, echo = TRUE,  fig.height = 5}
set.seed(2); x <- rnorm(100)

par(bg = "aliceblue") # Aliceblue background color

plot.new(); plot.window(xlim = range(x), ylim = c(0.5, 1.5))

rect(par("usr")[1], par("usr")[3], 
     par("usr")[2], par("usr")[4],
     col = "#f7f7f7") # Change the plot region color



boxplot(x, col = 4, horizontal = T ,add = T) # Create your plot 
 
```


# More Flexible Layouts: `layout` 1
```{r, echo = TRUE, class.source="bg-success", fig.height = 4}
layout(rbind(c(0,4,4,0),
             c(0,2,0,0),
             c(0,1,3,0),
             c(0,0,0,0)),
      height = c(lcm(2), lcm(2), 1, lcm(2)),
      width  = c(lcm(2), 1, lcm(2), lcm(1)))
layout.show(4)
box("outer", lty = "dotted")
```






# More Flexible Layouts: `layout` 2

\tiny
```{r, message=FALSE, echo =TRUE, fig.height= 4}
layout(rbind(c(0,4,4,0),
             c(0,2,0,0),
             c(0,1,3,0),
             c(0,0,0,0)),
     height = c(lcm(2), lcm(2), 1, lcm(2)),
      width = c(lcm(2), 1, lcm(2), lcm(1)))

par(mar = rep(0, 4), cex = 1.5, bg = 'lightblue') # setting the parameters
x = 1:30
y = rnorm(30) + x/5
plot(x+0.1, y-0.1, pch = 19, col = 'gray60'); points(x, y, pch = 19 , col = 'red') # First figure
boxplot(x, horizontal = TRUE, col = 'orange' ,axes = FALSE) # Second figure
boxplot(y, col = 'green', axes = FALSE) # Third figure
plot.new(); plot.window(xlim = c(0, 1), ylim = c(0, 1)) # Forth figure
text(.5, 0.25, "An Enhanced Scatterplot", cex = 1.5, font = 2)
```



```{r}
plot.new()
mydata = rnorm(50, 1, 4)
n = length(mydata)
seq(-5,5, length = n)
mymax = .15
plot.window(xlim = range(mydata) , ylim = c(0, mymax) )

hist(mydata, add = T, freq = F)
points(mydata, rep(.01, n), col = 2, pch = 19)
points(mydata, jitter(rep(.05, n), amount = .025), col = 3, pch = 19)
axis(1, at = c(-8,8))
axis(2)

```

# Trellis and Lattice

- R also provides an implementation of Trellis
plots via the package `lattice` by Deepayan Sarkar, on the basis of 
the “Grid” graphics system written by Paul Murrell of
Auckland.

- Trellis plots embody a number of design principles and these principles are evident in a number of new plot types in Trellis and in the default choice of colors, symbol shapes, and line styles provided by
Trellis plots.

- Trellis plots provide a feature known as multipanel conditioning, which creates multiple plots by splitting the data being plotted according to the levels of other variables.

# Why `lattice` system?

- The default appearance of the `lattice` plots is superior in some areas. The default colors and the default data symbols have been deliberately chosen to make it \textcolor{blue}{easy to distinguish}.

- The arrangement of plot components is \textcolor{blue}{more automated} in `lattice`. It is usually not necessary to set figure margins
manually.

- Legends can be automatically generated by the `lattice` system.

- The output from lattice functions is `grid` output, so many powerful `grid` features are \textcolor{blue}{available for annotating, editing, and saving the graphics output}.




# The `lattice` system structure in R 

 `graph_type(formula, data = )`
 

\small
formula    | description   
------------- | -------------
 `~ y`  |   Some univariate plot (boxplot, histogram, boxplot, …)
 `~ y | A`  |       Univariate, separate panels for levels of factor A
 `~ y | z`  |       Univariate, cutting z into discrete ranges
            |
 `y ~ x`    |       Bivariate
 `y ~ x | A` |         Bivariate, separate panels for levels of A
 `y ~ x | A + B` |    Multiple conditioning variables
 `y1 + y2 ~ x1 + x2` | Multiple Y and X variables

\color{blue}
graph_type    | description   | graph_type    | description  
------------- | ------------- | ------------- | -------------
`barchart`	|bar chart | `bwplot`	|boxplot	
`cloud`	| 3D scatterplot| `contourplot`	|3D contour plot	
`densityplot`|	kernal density plot	|`dotplot` |	dotplot	
`histogram`	| histogram	| `levelplot`	| 3D level plot	
`splom`	| scatterplot matrix	| `stripplot`	| strip plots
`xyplot` |	scatterplot	| `wireframe`	| 3D wireframe 	


# Types of the the `lattice` system
```{r,  message=FALSE}
library(lattice)
library(gridGraphics)
x <- 1:5
y <- 1:5
g <- factor(1:5)
types <- c("barchart", "bwplot", "densityplot", "dotplot",
           "histogram", "qqmath", "stripplot", "qq",
           "xyplot", "levelplot", "contourplot",
           "cloud", "wireframe", "splom", "parallelplot")
angle <- seq(0, 2*pi, length=19)[-19]
xx <- cos(angle)
yy <- sin(angle)
gg <- factor(rep(1:3, each=6))

aaa <- seq(0, pi, length=10)
xxx <- rep(aaa, 10)
yyy <- rep(aaa, each=10)
zzz <- sin(xxx) + sin(yyy)


doplot <- function(name, ...) {
  do.call(name, 
          list(..., scales=list(draw=FALSE), xlab=NULL, ylab=NULL,
               strip=function(which.panel, ...) { 
                       grid.rect(gp=gpar(fill="gray90")); grid.text(name) 
                     }))
}
plot <- vector("list", 15)
plot[[1]] <- doplot("barchart", y ~ g | 1)
plot[[2]] <- doplot("bwplot", yy ~ gg | 1, 
                    par.settings=list(box.umbrella=list(lwd=0.5)))
plot[[3]] <- doplot("densityplot", ~ yy | 1)
plot[[4]] <- doplot("dotplot", g ~ y | 1)
plot[[5]] <- doplot("histogram", ~ xx | 1)
plot[[6]] <- doplot("qqmath", ~ yy | 1)
plot[[7]] <- doplot("stripplot", yy ~ gg | 1)
plot[[8]] <- doplot("qq", gg ~ yy | rep(1, 18), subset=gg != 3)
plot[[9]] <- doplot("xyplot", xx ~ yy | 1)
plot[[10]] <- doplot("levelplot", zzz ~ xxx + yyy | 1, colorkey=FALSE)
plot[[11]] <- doplot("contourplot", zzz ~ xxx + yyy | 1, labels=FALSE, cuts=8)
plot[[12]] <- doplot("cloud", zzz ~ xxx + yyy | 1, zlab=NULL, zoom=0.9, 
                     par.settings=list(box.3d=list(lwd=0.1)))
plot[[13]] <- doplot("wireframe", zzz ~ xxx + yyy | 1, zlab=NULL, zoom=0.9,
                     drape=TRUE, par.settings=list(box.3d=list(lwd=0.1)),
                     colorkey=FALSE)
plot[[14]] <- doplot("splom", ~ data.frame(x=xx[1:10], y=yy[1:10]) | 1, 
                     pscales=0)
plot[[15]] <- doplot("parallelplot", ~ as.data.frame(split(yy, gg)) | 1)

grid.newpage()
pushViewport(viewport(layout=grid.layout(4, 4)))
for (i in 1:15) {
  pushViewport(viewport(layout.pos.col=((i - 1) %% 4) + 1,
                        layout.pos.row=((i - 1) %/% 4) + 1))
  print(plot[[i]], newpage=FALSE, 
        panel.width=list(1.025, "inches"),
        panel.height=list(1.025, "inches"))
  popViewport()
}
popViewport()
 

```

# Histogram and Box-Wishker plot in `lattice`
\scriptsize
::: columns

:::: column

```{r, echo=TRUE, fig.width= 8, fig.height = 6}

bwplot(age~pclass | survived, 
       data = Titanic )


```
:::: 

:::: column
```{r, echo=TRUE, fig.width = 8, fig.height = 7}
# Condition on Passenger Class & Gender
histogram(~ age | pclass, 
  data   = Titanic,
  layout = c(3,1),
   strip = strip.custom(
           strip.names = TRUE,
            var.name   = 
             "Passenger Class"),
    xlab = "Age")
```

::::

:::::




# Barchart 
\scriptsize
::: columns

:::: column

```{r, echo=TRUE, fig.width= 8, fig.height = 6}
surv.tab = xtabs(~survived + 
                  pclass + 
                  Gender, 
                  data = Titanic)
surv.df = as.data.frame(surv.tab)

# In barchart you need Freq ~ X
barchart(Freq ~ survived,
         data = surv.df,
        xlab = "survived")


```
:::: 

:::: column
```{r, echo=TRUE, fig.width = 8, fig.height = 7}
# Condition on Passenger Class & Gender
barchart(
  Freq ~ survived | pclass * Gender, 
  data   = surv.df, origin = 0,
   strip = strip.custom(
           strip.names = TRUE,
            var.name   = 
      c("Passenger Class", "Gender")),
    xlab = "Survived")
```

::::

:::::



::: columns

:::: column

# `xyplot`: Scatter plot in `lattice`
\small 
```{r, echo=T, fig.show='hold', fig.height=3.5, fig.width= 8.5}
library(lattice)
xyplot(lat ~ long | cut(depth, 3), 
       data = quakes, aspect = "iso", pch = ".", cex = 2,
       type = c("p", "g"), xlab = "Longitude", ylab="Latitude", 
       strip = strip.custom(strip.names = TRUE, 
                            var.name = "Depth"))
```

# `splom`: Scatter Plot Matrix
\small
```{r, echo=TRUE, fig.width= 6, fig.height = 4}
splom(~USArrests[c(3, 1, 2, 4)] | state.region, 
      pscales = 0, type = c("g", "p", "smooth"), 
      layout = c(2,2))
```

# panel function in lattice 
\scriptsize
::: columns

:::: column

```{r, echo=TRUE, fig.width= 8, fig.height = 6}
xyplot(mpg ~ disp | factor(gear), 
                  data = mtcars,
  layout=c(3, 1), aspect=1,
  
  panel = function(x, y, ...) {
  panel.lmline(x, y)
  panel.xyplot(x, y, ...)
})
```
:::: 

:::: column
```{r, echo=TRUE, fig.width = 8, fig.height = 6}
xyplot(mpg ~ disp | factor(gear), 
       data=mtcars,
layout=c(3, 1), aspect = 1,
panel = function(...) {
panel.xyplot(...)
panel.abline(h = 29, lty = "dashed")
panel.text(470,29.5, "efficiency",
           adj = c(1, 0), cex= 0.9)
})
```

::::

:::::






# Most common used `panel`s
\small 
A selection of predefined panel functions for adding graphical output to the panels of lattice plots.

\color{purple}
Function | Description
------------- | -------------
`panel.points()` | Draw data symbols at locations (x, y)
`panel.lines()` | Draw lines between locations (x, y)
`panel.segments()` | Draw line segments between (x0, y0) and (x1, y1)
`panel.arrows()` | Draw line segments and arrowheads to the end(s)
`panel.polygon()` | Draw one or more polygons with vertices (x, y)
`panel.text()` | Draw text at locations (x, y)
`panel.abline()` | Draw a line with intercept a and slope b
`panel.curve()` | Draw a function given by expr
`panel.rug()` | Draw axis ticks at x- or y-locations
`panel.grid()` |Draw a (gray) reference grid
`panel.loess()` | Draw a loess smooth through (x, y)
`panel.violin()` | Draw one or more violin plots
`panel.smoothScatter()` | Draw a smoothed 2D density of (x, y) 

# panel
\tiny
```{r, echo=TRUE, fig.width = 10, fig.height = 3.5}
depth.col = gray.colors(100)[cut(quakes$depth, 100, label = FALSE)]
depth.ord = rev(order(quakes$depth))
quakes$Magnitude=equal.count(quakes$mag,4); quakes$color=depth.col; quakes.ordered=quakes[depth.ord,]
depth.breaks = do.breaks(range(quakes.ordered$depth), 50)

xyplot(lat ~ long | Magnitude, data = quakes.ordered, 
       aspect = "iso", groups = color, cex = 2, col = "black",
       panel = function(x, y, groups, ..., subscripts) {
           fill <- groups[subscripts]
           panel.grid(h = -1, v = -1)
           panel.xyplot(x, y, pch = 21, fill = fill, ...)},
       
     legend = list(right = list(fun = draw.colorkey, args = 
                            list(key = list(col = gray.colors, at = depth.breaks)))),
       xlab = "Longitude", ylab = "Latitude")

```


# Grammar of Graphics: `ggplot2` 1
Leland Wilkinson's \textcolor{blue}{Grammar of Graphics} provides another completely different paradigm for producing statistical plots and this approach to plotting has
been implemented for R by Hadley Wickham's `ggplot2` package.

- One advantage of this package is that it makes it possible to create a very
wide variety of plots from a relatively small set of fundamental components.
- The ggplot2 package also has a feature called facetting, which is similar to
lattice's multipanel plots.

# Grammar of Graphics: `ggplot2` 2

## Every graph can be described as a combination of
\large
  - \textcolor{purple}{data}: a data frame: \textcolor{blue}{quantitative, categorical};
  
\large
  - \textcolor{purple}{aes}thetic: \textcolor{blue}{mapping} of variables into visual properties: \textcolor{blue}{x, y, size, color, ...}
  
\large
  - \textcolor{purple}{geom}etric objects (“geom”): \textcolor{blue}{points, lines, areas, arrows, …}
  
\large
  - \textcolor{purple}{layer}s: graph elements combined with \textcolor{red}{“+”}
  - \textcolor{purple}{coord}inate system (“coord”): \textcolor{blue}{Cartesian, polar, log, map,}

 
## And some more

  - \textcolor{purple}{stat}istical transformations (“stat”) -- data summaries: \textcolor{blue}{mean, sd, binning \& counting, ...}
  - \textcolor{purple}{scale}s: \textcolor{blue}{legends, axes}
  - \textcolor{purple}{position} adjustments: \textcolor{blue}{jitter, dodge, stack, …}
  - \textcolor{purple}{facet}ing: \textcolor{blue}{small multiples or conditioning to break a plot}

# `ggplot` 1 

```{r, error=TRUE, echo= TRUE, message=FALSE, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

```{r, echo= TRUE, fig.height=2.25, fig.width = 4.5}
ggplot(data = mpg) + 
  geom_point(mapping = 
               aes(x = displ, y = hwy, color = class))
```

# # `ggplot` 2  
```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE }
# Left
ggplot(data = mpg) + 
  geom_point(mapping = 
               aes(x = displ, y = hwy, alpha = class))
# Right
ggplot(data = mpg) + 
  geom_point(mapping = 
               aes(x = displ, y = hwy, shape = class))
```


::: columns

:::: column

```{r, echo = F,  fig.height = 3, fig.width=4, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = 
               aes(x = displ, y = hwy, alpha = class))
```

::::

:::: column


```{r, echo = F,  fig.height = 3, fig.width=4, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = 
               aes(x = displ, y = hwy, shape = class))
```

::::

:::


# `facet`

```{r, fig.show = 'hold', error=TRUE, echo= TRUE, fig.height = 3, fig.width=6}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

# `geom_bar` 1

\small

```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE }
#Left
ggplot(data = Titanic) + 
  geom_bar(mapping = aes(x = pclass, colour = pclass),
           fill    = 'red',
           lwd     = 2)

# Right
ggplot(data    = Titanic,
       mapping = aes(x= pclass, colour= Gender, fill= pclass))+ 
  geom_bar(lwd = 2, fill = "white", position = "dodge") +
  facet_wrap(~ survived)
# position can be 'dodge' and fill, 'stack' is the default for bars

```


::: columns

:::: column

```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = Titanic) + 
  geom_bar(mapping = aes(x = pclass, colour = pclass), fill = 'red', lwd = 2)
```

::::

:::: column


```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = Titanic, mapping = aes(x = pclass, colour = Gender, fill = pclass)) + 
  geom_bar( lwd = 2, fill = "white", position="dodge")+
  facet_wrap(~survived)
```

::::

:::





# `geom_bar` 2

\small

```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE }

dplot = ggplot(Titanic, aes(pclass, fill = survived)) + 
  xlab(NULL) + ylab(NULL) + theme(legend.position = "none")

# position stack is the default for bars, so `geom_bar()` 
# is equivalent to `geom_bar(position = "stack")`.
#Left
dplot + geom_bar() + 
  theme(plot.background = element_rect(fill = "lightblue"))
# Right
dplot + geom_bar(position = "fill") + 
        theme(legend.position = "right") # the default 
```


::: columns

:::: column

```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
dplot = ggplot(Titanic, aes(pclass, fill = survived)) + 
  xlab(NULL) + ylab(NULL) + theme(legend.position = "none")

dplot + geom_bar()

```

::::

:::: column


```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
dplot = ggplot(Titanic, aes(pclass, fill = survived)) + 
  xlab(NULL) + ylab(NULL) + theme(legend.position = "none")
# Right
dplot + geom_bar(position = "fill")+ theme(legend.position = "right") # the default 
  
```

::::

:::

# `geom_box`



```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE }
#Left
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot(col = 'red')
#Right
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot(col = 'red', fill = 'green') +
  coord_flip()
```


::: columns

:::: column

```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot(col = 'red')

```

::::

:::: column


```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot(col = 'red', fill = 'green') +
  coord_flip()
```

::::

:::

# map

```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE }
library(maps); library(tidyverse)

UK <- map_data("world") %>% filter(region=="UK")
#Left
ggplot(UK, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
#Right
ggplot(UK, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```


::: columns

:::: column

```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
library(maps)
library(tidyverse)
UK <- map_data("world") %>% filter(region=="UK")
ggplot(UK, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
```

::::

:::: column


```{r, echo = F,  fig.height = 3, fig.width=5, error=TRUE,  message=FALSE, warning=FALSE}
UK <- map_data("world") %>% filter(region=="UK")
ggplot(UK, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

::::

:::

# map 2

\small
```{r, fig.show = 'hide', error=TRUE, echo= TRUE, message=FALSE, eval=FALSE, size= 'tiny' }
# Get a data frame with longitude, latitude, and size of bubbles (a bubble = a city)
library(ggrepel)
UK_city <- world.cities %>% filter(country.etc == "UK")
ggplot(data = UK) +
  geom_polygon(aes(x=long, y = lat, group = group), 
                fill="grey", alpha=0.3) +
  geom_point( data = UK_city, aes(x=long, y=lat)) +
  theme_void() + ylim(50,59) + coord_map() 

# Second graphic with names of the 10 biggest cities
ggplot(data = UK) +
  geom_polygon( aes(x=long, y = lat, group = group), 
                fill="grey", alpha=0.3) +
  geom_point(UK_city, aes(x=long, y=lat, alpha=pop)) +
  geom_text_repel(UK_city %>% arrange(pop) %>% tail(10), 
                  aes(x=long, y=lat, label=name), size=5) +
  geom_point(UK_city %>% arrange(pop) %>% tail(10), 
             aes(x=long, y=lat), color="red", size=3) +
  theme_void() + ylim(50,59) + coord_map() +
  theme(legend.position="none")
```

# map 3




::: columns

:::: column

```{r, echo = F,  fig.height = 10, fig.width=8, error=TRUE,  message=FALSE, warning=FALSE}
library(ggrepel)
UK_city <- world.cities %>% filter(country.etc == "UK")
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point( data = UK_city, aes(x=long, y=lat)) +
  theme_void() + ylim(50,59) + coord_map()
```

::::

:::: column


```{r, echo = F,  fig.height = 10, fig.width=8, error=TRUE,  message=FALSE, warning=FALSE}
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point( data=UK_city, aes(x=long, y=lat, alpha=pop)) +
  geom_text_repel( data=UK_city %>% arrange(pop) %>% tail(10),
                   aes(x=long, y=lat, label=name), size = 6) +
  geom_point( data=UK_city %>% arrange(pop) %>% tail(10),
              aes(x=long, y=lat), color="red", size=3) +
  geom_point(data = UK_city %>% filter(name == 'Lancaster'),
              aes(x=long, y=lat), color="blue", size = 4) +
  geom_text_repel( data = UK_city %>% filter(name == 'Lancaster'),
                   aes(x=long, y=lat, label = name), size = 4) +
  theme_void() + ylim(50,59) + coord_map() +
  theme(legend.position="none")
```

::::

:::


# map 4
\tiny
```{r, echo = T,  fig.show='hide', error=TRUE,  message=FALSE, warning=FALSE, size= 'small'}
# virids package for the color palette
library(viridis)
 
# Left: use size and color
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point( data=UK_city, aes(x=long, y=lat, size=pop, color=pop)) +
  scale_size_continuous(range=c(1,12)) +
  scale_color_viridis(trans="log") +
  theme_void() + ylim(50,59) + coord_map() 
 
# Center: reorder your dataset first! Big cities appear later = on top
UK_city %>%
 arrange(pop) %>% 
 mutate( name=factor(name, unique(name))) %>% 
 ggplot() +
    geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
    geom_point( aes(x=long, y=lat, size=pop, color=pop), alpha=0.9) +
    scale_size_continuous(range=c(1,12)) +
    scale_color_viridis(trans="log") +
    theme_void() + ylim(50,59) + coord_map() + theme(legend.position="none")
 
# Right: just use arrange(desc(pop)) instead
UK_city %>%
 arrange(desc(pop)) %>% 
 mutate( name=factor(name, unique(name))) %>% 
 ggplot() +
    geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
    geom_point( aes(x=long, y=lat, size=pop, color=pop), alpha=0.9) +
    scale_size_continuous(range=c(1,12)) +
    scale_color_viridis(trans="log") +
    theme_void() + ylim(50,59) + coord_map() + theme(legend.position="none")
```

# map 4_2

:::::: {.columns}

::: {.column width="35%"}

```{r, echo = F,  fig.width=5, fig.height= 7, error=TRUE,  message=FALSE, warning=FALSE}
# virids package for the color palette
library(viridis)
 
# Left: use size and color
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point( data=UK_city, aes(x=long, y=lat, size=pop, color=pop)) +
  scale_size_continuous(range=c(1,12)) +
  scale_color_viridis(trans="log") +
  theme_void() + ylim(50,59) + coord_map() 
 
```

:::

::: {.column width="30%"}


```{r, echo = F,   fig.width=5, fig.height=8, error=TRUE,  message=FALSE, warning=FALSE}
# Center: reorder your dataset first! Big cities appear later = on top
UK_city %>%
 arrange(pop) %>% 
 mutate( name=factor(name, unique(name))) %>% 
 ggplot() +
    geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
    geom_point( aes(x=long, y=lat, size=pop, color=pop), alpha=0.9) +
    scale_size_continuous(range=c(1,12)) +
    scale_color_viridis(trans="log") +
    theme_void() + ylim(50,59) + coord_map() + theme(legend.position="none")
 
```

:::

::: {.column width="30%"}


```{r, echo = F,   fig.width=5, fig.height= 8, error=TRUE,  message=FALSE, warning=FALSE}
# Right: just use arrange(desc(pop)) instead
UK_city %>%
 arrange(desc(pop)) %>% 
 mutate( name=factor(name, unique(name))) %>% 
 ggplot() +
    geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
    geom_point( aes(x=long, y=lat, size=pop, color=pop), alpha=0.9) +
    scale_size_continuous(range=c(1,12)) +
    scale_color_viridis(trans="log") +
    theme_void() + ylim(50,59) + coord_map() + theme(legend.position="none")
```

::::


::::::

# Resourses
\small
- Paul Murrell, R Graphics, 3rd Ed 
  + [R Code:][ https://www.stat.auckland.ac.nz/~paul/RG3e/]
- Hadley Wickham, ggplot2: Elegant graphics for data analysis, 2nd Ed.
  [ggplot2](https://ggplot2.tidyverse.org/)
  
  <https://ggplot2.tidyverse.org/reference/index.html#plot-basics>
  
  
- Winston Chang, R Graphics Cookbook: Practical Recipes for Visualizing Data  [R Graphics Cookbook](http://www.cookbook-r.com/Graphs)

- Antony Unwin, Graphical Data Analysis with R
  - [R code:][http://www.gradaanwr.net/]

## Useful online resouses 
- [R CHARTS](https://r-charts.com/)
- [Topic in Computational Data Analysis and Graphics](https://www.stat.auckland.ac.nz/~ihaka/787/slides.html)
- [Lattice: Multivariate Data Visualization with R - Figures and Code](http://lmdvr.r-forge.r-project.org/figures/figures.html)
- [Data Visualization in R](https://www.datavis.ca/courses/RGraphics/index.html)
  
  
